The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-02-14 script/en/chapters/paper_and_pencil.tex
First version of CT Book ed 12. Most important point is the newly includes bitciphers chapter. Further things: Tried to reduce latex warnings amap. Changed wording from Sage to SageMath where visible...

The following chapter provides a broad overview of paper and pencil 
methods\footnote{%
The footnotes of this chapter describe how these methods can be
performed using the offline programs CrypTool 1 ({\bf CT1})\index{CT1},
CrypTool 2 ({\bf CT2})\index{CT2} and JCrypTool ({\bf JCT})\index{JCT}.
See the appendices
\ref{s:appendix-menu-overview-CT1},
\ref{s:appendix-template-overview-CT2}, and
\ref{s:appendix-function-overview-JCT}.\\
Many of the methods can also per performed within a browser, e.g. on the
the website CrypTool-Online ({\bf CTO}) (\url{http://www.cryptool-online.org})\index{CTO}. See the appendix \ref{s:appendix-function-overview-CTO}.\\
Additionally the last sub chapter (\ref{PaP_Sage_samples})
contains example code for the computer-algebra system {\bf Sage}\index{Sage}.
                 }
each with references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

\item {\bf Affine cipher}\footnote{
   Some according Sage samples are implemented at \ref{PaP_affine_cipher_sage-sample}.
   }%
   \index{Affine cipher}: This is a generalization of
   the shift cipher. A plaintext character is first substituted by another
   character and then the result is encrypted using the shift cipher.
   The name ``affine cipher'' was chosen because its encryption and
   decryption can be described as affine or linear function.

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\newpage
\section{Appendix: Examples using Sage}
\label{PaP_Sage_samples}
\index{Sage!Code examples}
\index{Sage}

In the following section some classic ciphers are implemented using the open source computer
algebra system \verb!Sage!\footnote{%
    A first introduction to the CAS Sage can be found in the appendix \ref{s:appendix-using-sage}.
                        }
. The code was tested with Sage version 5.3.
All ciphers are explained in chapter~\ref{Kapitel_PaperandPencil}
(``\nameref{Kapitel_PaperandPencil}'').

\noindent To make the sample code\footnote{%
     Further examples with Sage concerning classic crypto methods can be found e.g.:\\
     - at \url{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}\\
     - in the thesis of Minh Van Nguyen \cite{pp:Nguyen2009}
                                          }
of the ciphers easier to understand, we used the structure and the naming conventions
shown in the Figure \ref{naming-convention-in-Sage-cipher-samples} below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \begin{itemize}
    \item Encoding adapts the letters in the given plaintext P to the case defined in the
          given alphabet, and all non-alphabet characters are filtered out.
    \item Enciphering creates the ciphertext C.
  \end{itemize}
  \item Decryption also consists of two steps: deciphering and decoding.
  \begin{itemize}
    \item Decoding is only necessary if the symbols in the alphabet are not ASCII characters.
  \end{itemize}
\end{itemize}

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.5]{figures/notation-enc-dec_in-sage-scripts_en}
\caption{Structure and naming convention of the Sage cipher code examples} 
\label{naming-convention-in-Sage-cipher-samples}
\end{center}
\end{figure}

@@

Transposition ciphers are implemented in the Sage class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains the symbols used to build the space of our plaintext
and ciphertext. 
Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in the simple columns
transposition. This row vector is our key, and it specifies a permutation
of a plaintext.

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # given plaintext
sage: P   = "a b c d e f g h i j k l m n"
sage: # encryption key
sage: key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars, convert lower-case to upper-case)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in two steps)
sage: E   = T(key)
sage: C   = E(msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: D   = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
sage:
sage: # Representation of key and inverse key as permutations
sage: E
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: D
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple transposition by shifting (key and inverse key explicitly given)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14, code more variable
sage: keylen = 14
sage: shift = 2
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen)
sage:
sage: # construct the plaintext string from the first 14 letters of the alphabet plus blanks
sage: # plaintext   = "A B C D E F G H I J K L M N"
sage: A.gens()
(A, B, C, D, E, F, G, H, I, J, K, L, M, N, O, P, Q, R, S, T, U, V, W, X, Y, Z)
sage: P=''
sage: for i in range(keylen): P=P + " " + str(A.gen(i))
....:
sage: P
' A B C D E F G H I J K L M N'
sage:
sage: # encryption key
sage: # key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: key = [(i+shift).mod(keylen) + 1 for i in range(keylen)]; key
[3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in one step)
sage: C   = T.enciphering(key, msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: # keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) + 1 for i in range(keylen)]; keyInv
[13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: DC   = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMN
sage:
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
sage: # using the deciphering method requires to change the type of the variable key
sage: DC  = T.deciphering(T(key).key(), C); DC
ABCDEFGHIJKLMN
sage:
sage: # representation of key and inverse key as permutations
sage: T(key)
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(key).key()
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(keyInv)
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple transposition by shifting (key and inverse key constructed with ``range'')}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # Remark: Enciphering here requires, that the length of msg is a multiple of keylen
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,3,13,6,5,4,12,7)(11,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage: C   = T.enciphering(key, msg); C
BCMLDEAHIJNGFKPQAZRSOVWXBUTY
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
ssage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Just another way of decryption: Using "enciphering" with the inverse key
sage: keyInv = T.inverse_key(key); keyInv
(1,7,12,4,5,6,13,3,2)(11,14)
sage: DC     = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Test correctness of decryption
sage: msg == DC
True
\end{Verbatim}
\caption{Simple column transposition with randomly generated (permutation) key}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage: T.key_space()
Symmetric group of order 14! as a permutation group
sage: # Remark: The key space is not quite correct as also permutations shorter than keylen are counted.
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # enciphering in one and in two steps
sage: C   = T.enciphering(key, msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: enc = T(key); enc.key()
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: C = enc(msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: # deciphering
sage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
\end{Verbatim}
\caption{Simple column transposition (showing the size of the key space)}% the key\_space
\end{sagecode}

@@

Substitution cryptosystems are implemented in Sage in the class
\begin{center}
\verb!sage.crypto.classical.SubstitutionCryptosystem!
\end{center}

\noindent The following code sample uses Sage to construct a substitution
cipher with a random key. A random key can be generated using the
method \verb!random_key()! of the class
\texttt{SubstitutionCryp\-to\-system}. Different keys determine different
substitution ciphers. So each call to \verb!random_key()! returns
different results.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: # We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with randomly generated key}
\end{sagecode}

@@

The following example uses Sage to construct a Caesar cipher.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: # Next, we recover the plaintext.
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; key explicitly given, step-by-step approach)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: S
Substitution cryptosystem on Free alphabetic string monoid on A-Z
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: len(key)
26
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: # Next, we recover the plaintext.
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; substitution key generated)}
\end{sagecode}

@@

The shift cipher is implemented directly in the Sage class
\begin{center}
\verb!sage.crypto.classical.ShiftCryptosystem!
\end{center}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct Shift cipher directly
sage: shiftcipher = ShiftCryptosystem(AlphabeticStrings()); shiftcipher
Shift cryptosystem on Free alphabetic string monoid on A-Z
sage: P = shiftcipher.encoding("Shift me any number of positions."); P
SHIFTMEANYNUMBEROFPOSITIONS
sage: key = 12  # shift can be any integer number
sage: 
sage: # shift the plaintext by 12 positions to get the ciphertext
sage: C = shiftcipher.enciphering(key, P); C
ETURFYQMZKZGYNQDARBAEUFUAZE
sage: 
sage: # decrypt the ciphertext and ensure that it is the original plaintext
sage: DC = shiftcipher.deciphering(key, C); DC
SHIFTMEANYNUMBEROFPOSITIONS
sage: DC == P
True
\end{Verbatim}
\caption{A shift cipher over the upper-case letters of the English alphabet}
\label{paper_pencil:shift_cipher:Sage_example}
\end{sagecode}

@@

The affine cipher is implemented in the Sage class
\begin{center}
\verb!sage.crypto.classical.AffineCryptosystem!
\end{center}

@@

The Vigen{\`e}re cipher is implemented in the Sage class
\begin{center}
\verb!sage.crypto.classical.VigenereCryptosystem!
\end{center}

>>>>

The following chapter provides a broad overview of paper and pencil 
methods\footnote{%
The footnotes of this chapter describe how these methods can be
performed using the offline programs CrypTool 1 ({\bf CT1})\index{CT1},
CrypTool 2 ({\bf CT2})\index{CT2} and JCrypTool ({\bf JCT})\index{JCT}.
See the appendices
\ref{s:appendix-menu-overview-CT1},
\ref{s:appendix-template-overview-CT2}, and
\ref{s:appendix-function-overview-JCT}.\\
Many of the methods can also per performed within a browser, e.g. on the
the website CrypTool-Online ({\bf CTO}) (\url{http://www.cryptool-online.org})\index{CTO}. See the appendix \ref{s:appendix-function-overview-CTO}.\\
Additionally the last sub chapter (\ref{PaP_Sage_samples})
contains example code for the computer-algebra system {\bf SageMath}\index{SageMath}.
                 }
each with references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

\item {\bf Affine cipher}\footnote{
   Some according SageMath samples are implemented at \ref{PaP_affine_cipher_sage-sample}.
   }%
   \index{Affine cipher}: This is a generalization of
   the shift cipher. A plaintext character is first substituted by another
   character and then the result is encrypted using the shift cipher.
   The name ``affine cipher'' was chosen because its encryption and
   decryption can be described as affine or linear function.

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\newpage
\section{Appendix: Examples using SageMath}
\label{PaP_Sage_samples}
\index{SageMath!Code examples}
\index{SageMath}

In the following section some classic ciphers are implemented using the open source computer
algebra system \verb!SageMath!\footnote{%
    A first introduction to the CAS SageMath can be found in the appendix \ref{s:appendix-using-sage}.
                        }
. The code was tested with SageMath version 5.3.
All ciphers are explained in chapter~\ref{Kapitel_PaperandPencil}
(``\nameref{Kapitel_PaperandPencil}'').

\noindent To make the sample code\footnote{%
     Further examples with SageMath concerning classic crypto methods can be found e.g.:\\
     - at \url{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}\\
     - in the thesis of Minh Van Nguyen \cite{pp:Nguyen2009}
                                          }
of the ciphers easier to understand, we used the structure and the naming conventions
shown in the Figure \ref{naming-convention-in-Sage-cipher-samples} below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \begin{itemize}
    \item Encoding adapts the letters in the given plaintext P to the case defined in the
          given alphabet, and all non-alphabet characters are filtered out.
    \item Enciphering creates the ciphertext C.
  \end{itemize}
  \item Decryption also consists of two steps: deciphering and decoding.
  \begin{itemize}
    \item Decoding is only necessary if the symbols in the alphabet are not ASCII characters.
  \end{itemize}
\end{itemize}

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.5]{figures/notation-enc-dec_in-sage-scripts_en}
\caption{Structure and naming convention of the SageMath cipher code examples} 
\label{naming-convention-in-Sage-cipher-samples}
\end{center}
\end{figure}

@@

Transposition ciphers are implemented in the SageMath class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains the symbols used to build the space of our plaintext
and ciphertext. 
Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in the simple columns
transposition. This row vector is our key, and it specifies a permutation
of a plaintext.

@@

% Using [fontsize=\footnotesize,fontshape=tt] caused:
%   LaTeX Font Warning: Font shape `T1/cmtt/m/tt' undefined
%   (Font)              using `T1/cmtt/m/n' instead on input line 730.
% so deleted the fontshape parameter.
\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
sage: # transposition cipher using a block length of 14
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # given plaintext
sage: P   = "a b c d e f g h i j k l m n"
sage: # encryption key
sage: key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars, convert lower-case to upper-case)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in two steps)
sage: E   = T(key)
sage: C   = E(msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: D   = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
sage:
sage: # Representation of key and inverse key as permutations
sage: E
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: D
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple transposition by shifting (key and inverse key explicitly given)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
sage: # transposition cipher using a block length of 14, code more variable
sage: keylen = 14
sage: shift = 2
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen)
sage:
sage: # construct the plaintext string from the first 14 letters of the alphabet plus blanks
sage: # plaintext   = "A B C D E F G H I J K L M N"
sage: A.gens()
(A, B, C, D, E, F, G, H, I, J, K, L, M, N, O, P, Q, R, S, T, U, V, W, X, Y, Z)
sage: P=''
sage: for i in range(keylen): P=P + " " + str(A.gen(i))
....:
sage: P
' A B C D E F G H I J K L M N'
sage:
sage: # encryption key
sage: # key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: key = [(i+shift).mod(keylen) + 1 for i in range(keylen)]; key
[3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in one step)
sage: C   = T.enciphering(key, msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: # keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) + 1 for i in range(keylen)]; keyInv
[13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: DC   = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMN
sage:
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
sage: # using the deciphering method requires to change the type of the variable key
sage: DC  = T.deciphering(T(key).key(), C); DC
ABCDEFGHIJKLMN
sage:
sage: # representation of key and inverse key as permutations
sage: T(key)
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(key).key()
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(keyInv)
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple transposition by shifting (key and inverse key constructed with ``range'')}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
sage: # Remark: Enciphering here requires, that the length of msg is a multiple of keylen
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,3,13,6,5,4,12,7)(11,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage: C   = T.enciphering(key, msg); C
BCMLDEAHIJNGFKPQAZRSOVWXBUTY
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
ssage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Just another way of decryption: Using "enciphering" with the inverse key
sage: keyInv = T.inverse_key(key); keyInv
(1,7,12,4,5,6,13,3,2)(11,14)
sage: DC     = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Test correctness of decryption
sage: msg == DC
True
\end{Verbatim}
\caption{Simple column transposition with randomly generated (permutation) key}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage: T.key_space()
Symmetric group of order 14! as a permutation group
sage: # Remark: The key space is not quite correct as also permutations shorter than keylen are counted.
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # enciphering in one and in two steps
sage: C   = T.enciphering(key, msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: enc = T(key); enc.key()
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: C = enc(msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: # deciphering
sage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
\end{Verbatim}
\caption{Simple column transposition (showing the size of the key space)}% the key\_space
\end{sagecode}

@@

Substitution cryptosystems are implemented in SageMath in the class
\begin{center}
\verb!sage.crypto.classical.SubstitutionCryptosystem!
\end{center}

\noindent The following code sample uses SageMath to construct a substitution
cipher with a random key. A random key can be generated using the
method \verb!random_key()! of the class
\texttt{SubstitutionCryp\-to\-system}. Different keys determine different
substitution ciphers. So each call to \verb!random_key()! returns
different results.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: # We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic substitution with randomly generated key}
\end{sagecode}

@@

The following example uses SageMath to construct a Caesar cipher.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: # Next, we recover the plaintext.
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; key explicitly given, step-by-step approach)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: S
Substitution cryptosystem on Free alphabetic string monoid on A-Z
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: len(key)
26
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: # Next, we recover the plaintext.
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; substitution key generated)}
\end{sagecode}

@@

The shift cipher is implemented directly in the SageMath class
\begin{center}
\verb!sage.crypto.classical.ShiftCryptosystem!
\end{center}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
sage: # construct Shift cipher directly
sage: shiftcipher = ShiftCryptosystem(AlphabeticStrings()); shiftcipher
Shift cryptosystem on Free alphabetic string monoid on A-Z
sage: P = shiftcipher.encoding("Shift me any number of positions."); P
SHIFTMEANYNUMBEROFPOSITIONS
sage: key = 12  # shift can be any integer number
sage: 
sage: # shift the plaintext by 12 positions to get the ciphertext
sage: C = shiftcipher.enciphering(key, P); C
ETURFYQMZKZGYNQDARBAEUFUAZE
sage: 
sage: # decrypt the ciphertext and ensure that it is the original plaintext
sage: DC = shiftcipher.deciphering(key, C); DC
SHIFTMEANYNUMBEROFPOSITIONS
sage: DC == P
True
\end{Verbatim}
\caption{A shift cipher over the upper-case letters of the English alphabet}
\label{paper_pencil:shift_cipher:Sage_example}
\end{sagecode}

@@

The affine cipher is implemented in the SageMath class
\begin{center}
\verb!sage.crypto.classical.AffineCryptosystem!
\end{center}

@@

The Vigen{\`e}re cipher is implemented in the SageMath class
\begin{center}
\verb!sage.crypto.classical.VigenereCryptosystem!
\end{center}
